No one-size-fits-all solution to clean GBIF

### 2020-06-04 manuscript_peerj.Rmd
send to coauthors

    affiliation: 9

@@

    affiliation: 10,11,12

@@

    address: Universidade Estadual de Feira de Santana, Feira de Santana, Brazil

@@

    address: Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil

@@

  - code: 12

@@

  Species occurrence records provide the basis for many biodiversity studies. They derive from geo-referenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications. Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern. Automatic flagging and filtering are a scalable and reproducible means to identify potentially problematic records in datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org). However, it is unclear how much data may be lost by filtering, whether the same tests should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses. Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 18 Neotropical taxa including animals, fungi, and plants, terrestrial and marine, downloaded from GBIF. We find that 29-90% of the records are potentially erroneous, with large variation across taxonomic groups. Tests for duplicated information, collection year, basis of record as well as urban areas and coordinates for terrestrial taxa in the sea or marine taxa on land have the greatest effect. While many flagged records might not be *de facto* erroneous, they could be overly imprecise and increase uncertainty in downstream analyses. Automated flagging can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus. Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research. 

@@

@@

@@

# Introduction {-}
Publicly available species distribution data have become a crucial resource in biodiversity research, including studies in ecology, biogeography, systematics and conservation biology. In particular, the availability of digitized collections from museums and herbaria, and citizen science observations has increased drastically over the last few years. As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.3 billion geo-referenced occurrence records for species from across the globe and the tree of life.

A central challenge to the use of these publicly available species occurrence data in research are erroneous geographic coordinates [@Anderson2016]. Errors mostly arise because public databases integrate records collected with different methodologies in different places, at different times; often without centralized curation and only rudimentary meta-data. For instance, erroneous coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common [@Maldonado2015; @Yesson2007] and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in captivity [@Zizka2019].

Manual data cleaning based on expert knowledge can detect these issues, but it is only feasible on small taxonomic or geographic scales, and it is time-consuming and difficult to reproduce. As an alternative, automated flagging methods to identify potentially problematic records have been proposed as a scalable option, as they are able to deal with datasets containing up to millions of records and many different taxa. Those methods are usually based on geographic gazetteers [e.g., @Chamberlain2016; @Zizka2019; @Jin2020] or on additional data, such as environmental variables [@Robertson2016]. Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the suitability of publicly available occurrence records for biodiversity research [@Zizka2019].

Erroneous records are especially problematic in conservation, where stakes are high. Recently proposed methods for automated conservation assessments could support the formal assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) [@Dauby2017; @Bachman2011; @Pelletier2018]. These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B ("Geographic range"). These methods have been used to propose preliminary global [@Stevart2019; @Zizka2019a] and regional [@Schmidt2017; @Cosiaux2018] Red List assessments. However, all metrics but especially EOO, are sensitive to individual records with erroneous coordinates. Automated conservation assessments may therefore be biased, particularly if the number of records are few, as it is the case for many tropical species.

While newly proposed methods for automated cleaning of species records hold great promise for biodiversity research, their use across taxonomic groups and datasets remains poorly explored. Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different groups of animals, fungi, and plants. Furthermore, we test the impact of automated filtering procedures for the accuracy of automated conservation assessments compared to full IUCN assessments. Specifically, we evaluate a pipeline of 13 automated filters to flag possibly erroneous records by using 1) record meta-data and 2) geographic gazetteers. We address three questions:

@@

We downloaded occurrence information for all study groups from www.gbif.org using the `rgbif` v1.4.0 package [@Chamberlain2017] in R (GBIF.org, 2019a-p,2020a,b). We downloaded only records with geographic coordinates and limited the study area to latitudes smaller than 33$^\circ$ N and longitudes smaller than 35$^\circ$ W and larger than 120$^\circ$ W reflecting the Neotropics, our main area of expertise [@Morrone2014]. The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, Entomobryidae, *Gaylussacia*, Iridaceae, Neanuridae, and *Tillandsia* for which we only obtained the Neotropical occurrences.

##  Automated cleaning {-}
We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99. Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e., test for duplicated records). In a second step, we used the `clean_coordinates` function of the `CoordinateCleaner v2.0-11` package [@Zizka2019] with default options to flag errors that are common to biological data sets (“filtered”, hereafter). These include: coordinates in the sea for terrestrial taxa and on land for marine taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions. See Table 2 for a summary of all flagging tests we used.

## Downstream analyses {-}
We first generated species richness maps using 100x100 km grid cells for the raw and filtered datasets respectively, using the package `speciesgeocodeR v2.0-10` [@Topel2016]. We then performed an automated conservation assessment for all study groups based on both datasets using the `ConR v1.2.4` package [@Dauby2017]. `ConR` estimates the EOO, AOO, and the number of locations, and then suggests a preliminary conservation status based on Criterion B of the global IUCN Red List. While these assessments are preliminary [see @IUCN2017], they can be a proxy used by the IUCN to speed up full assessments. We then benchmarked the preliminary conservation assessments against the global IUCN Red List assessments as far as available for the same taxa, which we obtained from www.iucn.org via the `rredlist v.0.5.0` package [@Chamberlain2018].

## Evaluation of results {-}
<!-- This session is exciting and makes the study more reliable and interesting than one just based on data mining, but how this was done is very unclear. Did people provide a specific score for those, or how were their assessments recorded? I think we need to clarify this, as it’s the weakest part of the paper at the moment. We don’t report the results for this evaluation nor what they mean.  -->
Each author provided an informed comment on the performance of the raw and cleaned datasets, respectively concerning the number of flagged records and the accuracy of the overall species richness maps and we compared the agreement between automated conservation assessments based on raw and filtered occurrences with the global IUCN Red List (www.iucn.org).

We did all analyses in the R computing environment [@rcoreteam2019], using standard libraries for data handling and visualization [@Wickham2018; @Garnier2018; @Ooms2014; @Ooms2019; @Hijmans2019]. All scripts are available from a zenodo repository (**DOI**).

@@

tab1 <- tab1 %>%
  mutate_all(linebreak) %>%
  select(-researcher) %>% 
  filter(!taxon %in% c("Opisthonema", "Eugenia"))

@@

tab2 <- tab2[,c(1:3,16,5,7,4,8,6,14,10,15,11:13,9)]

tab2[,-c(1:4)] <- round((tab2[,-c(1:4)] / unlist(tab2[,2]) * 100), 1)

tab2 <- tab1 %>% 
  select(taxon) %>%  
  bind_rows(data.frame(taxon = "Total")) %>% 
  left_join(tab2, by = "taxon") %>% 
  replace_na(list(total_records = 0))

@@

tab3 <- tab1 %>% 
  select(taxon) %>%  
  left_join(tab3, by = "taxon") %>% 
  replace_na(list(iucn_spnum = 0, raw_spnum = 0, flag_spnum = 0)) %>% 
  filter(iucn_spnum > 1)

@@

We retrieved a total of `r nrow(dat)` species occurrence records, with a median of `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per study group and `r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per species (Table 3, Appendix 1). We obtained most records for `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(taxon) %>%  unlist()` (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs) %>%  unlist()`) and fewest for *`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(taxon) %>%  unlist()`* (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(nrecs) %>%  unlist()`). The species with most records was *`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(species)`* (`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs)`).

Our automated tests flagged a total of `r sum(!dat$summary)` records (Fig. \ref{fig:total}), with a median of `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per group. Overall, the most important test was for duplicated records (on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!.dpl)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per taxonomic group). The filtering steps based on record meta-data that flagged the largest number of records were the basis of records (`r dat %>% group_by(taxon) %>% summarize(num = mean(!coordinate_base)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the collection year (`r dat %>% group_by(taxon) %>% summarize(num = mean(!record_age)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%). The most important automated tests were for urban area (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.urb)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the occurrence from records of terrestrial taxa in the sea and marine taxa on land (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.sea)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%, see Table 3 and Appendix 1 in the electronic supplement for further details and the absolute numbers).  Only few records were flagged by the coordinate precision, zero coordinates and biodiversity institution tests (Fig. \ref{fig:split}).

*`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(1) %>% unlist()`*, *`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(2) %>% unlist()`*, and *`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(3) %>% unlist()`* had the highest fraction of flagged records (Table 3). In general, the different filters we tested were of similar importance for different study groups. There were few outstanding exceptions, including the particularly high proportions of records flagged on the "basis of record test" for *Tityus* (7.0%), Dipsadidae (5.6%), *Prosthechea* (5.0%) and *Tillandsia* (4.9%), by the collection year for Dipsadidae (11.3%), by the taxonomic identification level for  *Tityus* (1.6%), by the capital coordinates for *Oocephalus* (6.1%) and *Gaylussacia* (3.2%), by the seas/land test for Diogenidae and *Thozetella*, and by the urban areas test for *Oocephalus* (13.3%) and Iridaceae (12.3%). Furthermore, Entomobryidae differed considerably from all other study taxa with exceptionally high numbers of records flagged by the "basis of record, "level of identification' and urban areas" tests.

Geographically, the records flagged by the "basis of record" and "individual count" tests were concentrated in Central America and southern North America, and a relatively high number of records flagged due to their proximity to the centroids of political entities were located on Caribbean islands (Fig. \ref{fig:split}). See Appendix 2 for species richness maps using the raw and cleaned data for all study groups.

@@

# Discussion {-}
<!-- add some subheadings to discussion -->
Automated flagging based on meta-data and automatic tests flagged on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group. The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (Question 1). The importance of different tests was similar across taxonomic groups, with exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (Question 2). The results for species richness were similar between the raw and filtered data with some improvements by the filters. We found little impact of filtering on the accuracy of the automated conservation assessments (Question 3).

## The relevance of individual filters {-}
Several reasons may explain the high proportion of records flagged as duplicates. First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a specimen duplication is easily feasible. Second, independent collections at similar localities may occur, in particular for local endemics. Third, low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities. Duplicated records do not represent errors *per se*, but an unnecessary computational burden, which is often a bottleneck for analyses of datasets with millions of records.

The aim of automated filtering is to identify possibly problematic records that are unsuitable for particular downstream analyses. Hence, our tests may identify true errors (e.g., wrong coordinates such as those with exclusively zeros, equal latitude and longitude and records of terrestrial species in the sea or marine species on land) as well as records that are not necessarily erroneous, but have a low precision (e.g., collection year, coordinate precision, centroids of political units) or are redundant for certain analyses (duplicates). The importance of each test and the linked thresholds must be judged based on the specific downstream analyses. For instance, records assigned to country centroids might be acceptable for inter-continental comparisons, but can be considered erroneous for species distribution modelling on a local scale. As our results show, adapting automated tests to the geographic study area or the taxonomic study group can be warranted. For instance, the high number of records flagged for centroids on the Lesser Antilles (Fig. \ref{fig:split}) might be overly strict, although we chose a conservative distance for the Political centroid test (1 km).

## Similarities and differences among taxa {-}
The number of records flagged by individual tests was similar across study groups, suggesting that similar error types might be relevant for collections of plants and animals. Therefore, the same tests can be used across taxonomic groups. Some notable exceptions stress the need to adapt each test to the taxonomic study group to balance data quality and data availability. The high fraction of records flagged by the "basis of record test" for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups. The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection effort already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700. The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700. The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae, and *Oocephalus* might be related to horticulture since at least some species in those groups are commonly cultivated as ornamentals.

@@

Our results stress the importance of (meta-)data exploration prior to most biodiversity analyses. Automated flagging can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest. The putative errors we encountered point to the importance to train researchers and students to curate species occurrence datasets and to allow users to provide feedback for particular records, thus contributing to overall accuracy and precision of public biodiversity databases.

>>>>

@@

    affiliation: 9,10,11

@@

    address: Departamento de Botânica, Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil

@@

@@

@@

  Species occurrence records provide the basis for many biodiversity studies. They derive from geo-referenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications. Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern. Automatic filtering is a scalable and reproducible means to identify potentially problematic records and tailor datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org), for biodiversity analyses. However, it is unclear how much data may be lost by filtering, whether the same filters should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses. Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 18 Neotropical taxa including animals, fungi, and plants, terrestrial and marine, downloaded from GBIF. We find that a total of 44.3% of the records are potentially problematic, with large variation across taxonomic groups (25 - 90%). A small fraction of records was identified as erroneous in the strict sense (4.2%), the majority of filtered records that were rather unfit for downstream analyses (41.7%). Filters of duplicated information, collection year, basis of record as well as urban areas and coordinates for terrestrial taxa in the sea or marine taxa on land have the greatest effect. Automated filtering can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus. Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research.

@@

header-includes:
  - \usepackage[export]{adjustbox}

@@

<!-- # Keywords -->
<!-- Automated cleaning, automated conservation assessment, data quality, GBIF, Neotropics. -->

@@

# Introduction {-}
Publicly available species distribution data have become a crucial resource in biodiversity research, including studies in ecology, biogeography, systematics and conservation biology. In particular, the availability of digitized collections from museums and herbaria, and citizen science observations has increased drastically over the last few years. As of today, the largest public aggregator for geo-referenced species occurrences data, the Global Biodiversity Information Facility (www.gbif.org), provides access to more than 1.4 billion geo-referenced occurrence records for species from across the globe and the tree of life.

A central challenge to the use of these publicly available species occurrence data in research are problematic geographic coordinates, which are either erroneous or unfit for downstream analyses [for instance because they are overly imprecise, @Anderson2016]. Problems mostly arise because public databases integrate records collected with different methodologies in different places, at different times; often without centralized curation and only rudimentary meta-data. For instance, problematic coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common [@Maldonado2015; @Yesson2007] and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in captivity [@Zizka2019].

Manual data cleaning based on expert knowledge can detect these issues, but it is only feasible on small taxonomic or geographic scales, and it is time-consuming and difficult to reproduce. As an alternative, automated filtering methods to identify potentially problematic records have been proposed as a scalable option, as they are able to deal with datasets containing up to millions of records and many different taxa. Those methods are usually based on geographic gazetteers [e.g., @Chamberlain2016; @Zizka2019; @Jin2020] or on additional data, such as environmental variables [@Robertson2016]. Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the suitability of publicly available occurrence records for biodiversity research [@Zizka2019].

Problematic records are especially critical in conservation, where stakes are high. Recently proposed methods for automated conservation assessments could support the formal assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) [@Dauby2017; @Bachman2011; @Pelletier2018]. These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B ("Geographic range"). These methods have been used to propose preliminary global [@Stevart2019; @Zizka2020b] and regional [@Schmidt2017; @Cosiaux2018] Red List assessments. However, all metrics and especially EOO, are sensitive to individual records with problematic coordinates. Automated conservation assessments may therefore be biased, particularly if the number of records are few, as it is the case for many tropical species.

While automated filters hold great promise for biodiversity research, their use across taxonomic groups and datasets remains poorly explored. Here, we test the effect of automated filtering of species geographic occurrence records on the number of records available in different groups of animals, fungi, and plants. Furthermore, we test the impact of automated filtering procedures for the accuracy of preliminary automated conservation assessments compared to full IUCN assessments. Specifically, we evaluate a pipeline of 13 automated filters to flag possibly problematic records by using record meta-data and geographic gazetteers in two categories: 1) erroneous (coordinates, that are likely wrong, irrespective of the downstream analyses, for instance due to data entry errors) and 2) unfit for purpose (coordinates that are not wrong *per se*, but likely unfit for the planned downstream analyses, for instance because they are overly imprecise). We address three questions:

@@

We downloaded occurrence information for all study groups from www.gbif.org using the `rgbif` v1.4.0 package [@Chamberlain2017] in R (GBIF.org, 2019a-p,2020a,b). We downloaded GBIF-interpreted data including only records with geographic coordinates and limited the study area to a rectangle between 90$^\circ$S - 33$^\circ$ N and 35$^\circ$ W - 120$^\circ$ W reflecting the Neotropics [@Morrone2014], our main area of expertise. The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, Entomobryidae, *Gaylussacia*, Iridaceae, Neanuridae, and *Tillandsia* for which we only obtained the Neotropical occurrences. We consider GBIF data generally of high quality and use them as study case because GBIF is the largest, most widely-used and taxonomically most comprehensive data source for species occurrence records, but many more exist [e.g., https://bien.nceas.ucsb.edu/bien/, www.fishbase.de or @Guedes2018]. The variety of existing databases is also the reason why we download data without accounting for GBIF-internal quality testing ("Geospatial issues"), since quality control mechanisms might differ among databases and sometimes no quality control might be done (for instance in *de novo* databases compiled by groups of researchers).

##  Automated cleaning {-}
We followed the cleaning pipeline outlined by @Zizka2019 and first filtered the raw data downloaded from GBIF ("raw", hereafter) using meta-data for those records for which they were available [although this is often not the case, @Peterson2018], removing: 1) records with a coordinate precision below 100 km (as this represents the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99. Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e., test for duplicated records). In a second step, we used the `clean_coordinates` function of the `CoordinateCleaner v2.0-11` package [@Zizka2019] with default options to flag errors that are common to biological data sets (“filtered”, hereafter). These include: coordinates in the sea for terrestrial taxa and on land for marine taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions. See Table 2 for a summary of all filters we used and their classification into "erroneous" and "unfit".

## Downstream analyses {-}
We first generated species richness maps using 100x100 km grid cells for the raw and filtered datasets respectively, using the package `speciesgeocodeR v2.0-10` [@Topel2017]. We then performed an automated conservation assessment for all study groups based on both datasets using the `ConR v1.2.4` package [@Dauby2017]. `ConR` estimates the EOO, AOO, and the number of locations, and then suggests a preliminary conservation status based on Criterion B of the global IUCN Red List. While these assessments are preliminary [see @IUCN2017], they can be a proxy used by the IUCN to speed up full assessments. We then benchmarked the preliminary conservation assessments against the global IUCN Red List assessments as far as available for the same taxa, which we obtained from www.iucn.org via the `rredlist v.0.5.0` package [@Chamberlain2018].

## Evaluation of results {-}
Each author provided an informed comment on the performance of the raw and cleaned datasets, concerning the number of removed records and the accuracy of the overall species richness maps. We then compared the agreement between automated conservation assessments based on raw and filtered occurrences with the global IUCN Red List for those taxa where IUCN assessments were available (www.iucn.org).

We did all analyses in the R computing environment [@rcoreteam2019], using standard libraries for data handling and visualization [@Wickham2018; @Garnier2018; @Ooms2014; @Ooms2019; @Hijmans2019]. All scripts are available from a zenodo repository (doi:10.5281/zenodo.3695102).

@@

tab1 <- tab1 %>%
  mutate_all(linebreak) %>%
  dplyr::select(-researcher) %>% 
  filter(!taxon %in% c("Opisthonema", "Eugenia"))

@@

tab2 <- tab2[,c(1:3,16:17,14,12,9,10,15,11,13,5,7,4,8,6,18)]

err <- tab2[19,4] <-  base::colSums(tab2[, 4], na.rm = TRUE)
unfit <- tab2[19,5] <-  base::colSums(tab2[, 5], na.rm = TRUE)

tab2[,-c(1:2,18)] <- round((tab2[,-c(1:2,18)] / unlist(tab2[,2]) * 100), 1)

tab2 <- tab1 %>% 
  dplyr::select(taxon) %>%  
  bind_rows(data.frame(taxon = "Total")) %>% 
  left_join(tab2, by = "taxon") %>% 
  replace_na(list(total_records = 0))

@@

tab3 <- tab1 %>% 
  dplyr::select(taxon) %>%  
  left_join(tab3, by = "taxon") %>% 
  replace_na(list(iucn_spnum = 0, raw_spnum = 0, flag_spnum = 0)) %>% 
  filter(iucn_spnum > 1)

@@

We retrieved a total of `r nrow(dat)` species occurrence records, with a median of `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist() %>%  round(0)` records per study group and `r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% ungroup() %>%  summarize(median(nrecs)) %>%  unlist()` records per species (Table 3, Appendix 1). We obtained most records for `r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(taxon) %>%  unlist()` (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs) %>%  unlist()`) and fewest for *`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(taxon) %>%  unlist()`* (`r dat %>% group_by(taxon) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(-1) %>% select(nrecs) %>%  unlist()`). The species with most records was *`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(species)`* (`r dat %>% group_by(species) %>% dplyr::summarize(nrecs = n()) %>% arrange(nrecs) %>%  top_n(1) %>% select(nrecs)`).

Our automated tests filtered a total of `r sum(!dat$summary)` records (Fig. \ref{fig:total}, erroneous: `r err`, unfit: `r unfit`), with a median of `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per group (erroneous: `r median(tab2$total_error)`%, unfit: `r median(tab2$total_unfit)`%). Overall, the most important test was for duplicated records (on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!.dpl)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% per taxonomic group). The filtering steps based on record meta-data that filtered the largest number of records were the basis of records (`r dat %>% group_by(taxon) %>% summarize(num = mean(!coordinate_base)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the collection year (`r dat %>% group_by(taxon) %>% summarize(num = mean(!record_age)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%). The most important automated tests were for urban area (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.urb)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%) and the occurrence from records of terrestrial taxa in the sea and marine taxa on land (`r dat %>% group_by(taxon) %>% summarize(num = mean(!.sea)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`%, see Table 3 and Appendix 1 in the electronic supplement for further details and the absolute numbers).  Only a few records were filtered by the coordinate precision, zero coordinates and biodiversity institution tests (Fig. \ref{fig:split}).

`r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(1) %>% unlist()`, `r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(2) %>% unlist()`, and `r tab2 %>% arrange(desc(fraction)) %>% top_n(3) %>% select(taxon) %>% slice(3) %>% unlist()` had the highest fraction of filtered records (Table 3). In general, the different filters we tested were of similar importance for different study groups. There were few outstanding exceptions, including the particularly high proportions of records filtered on the "basis of record test" for *Tityus* (7.0%), Dipsadidae (5.6%), *Prosthechea* (5.0%) and *Tillandsia* (4.9%), by the collection year for Dipsadidae (11.3%), by the taxonomic identification level for  *Tityus* (1.6%), by the capital coordinates for *Oocephalus* (6.1%) and *Gaylussacia* (3.2%), by the seas/land test for Diogenidae and *Thozetella*, and by the urban areas test for *Oocephalus* (13.3%) and Iridaceae (12.3%). Furthermore, Entomobryidae differed considerably from all other study taxa with exceptionally high numbers of records filtered by the "basis of record, "level of identification' and urban areas" tests.

Geographically, the records filtered by the "basis of record" and "individual count" tests were concentrated in Central America and southern North America, and a relatively high number of records filtered due to their proximity to the centroids of political entities were located on Caribbean islands (Fig. \ref{fig:split}). See Appendix 2 for species richness maps using the raw and cleaned data for all study groups.

@@

# Discussion {-}
<!-- add some subheadings to discussion -->
Automated flagging based on meta-data and automatic tests filtered on average `r dat %>% group_by(taxon) %>% summarize(num = mean(!summary)) %>%  select(num) %>% unlist() %>%  mean() %>% round(3) * 100`% of the records per taxonomic group; `r min(tab2$total_unfit)`% - `r max(tab2$total_unfit)`% as "unfit" and `r min(tab2$total_error)`% - `r max(tab2$total_error)`% as "erroneous". The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (**Question 1**). The importance of different tests was similar across taxonomic groups, with exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (**Question 2**). The results for species richness were similar between the raw and filtered data with some improvements by the filters. We found little impact of filtering on the accuracy of the automated conservation assessments (**Question 3**).

## The relevance of individual filters {-}
The aim of automated filtering is to identify possibly problematic records that are unsuitable for particular downstream analyses. While those records filtered as "erroneous" will likely cause problems for most biodiversity research, those filtered as "unfit" might have varying impact, depending on the type and spatial resolution of the downstream analyses, ranging from an unnecessary computational burden which can be a bottle neck for large-scale analyses [i.e. duplicates, @Antonelli2018] over increasing uncertainty (due to low precision) to compromising results. For instance, records assigned to country centroids might be acceptable for inter-continental comparisons, but likely are erroneous for species distribution modelling on a local scale. The importance of each test and the linked thresholds must be judged based on the specific downstream analyses. As our results show, adapting automated tests to the geographic study area or the taxonomic study group can be warranted. For instance, the high number of records flagged for centroids on the Lesser Antilles (Fig. \ref{fig:split}) might be overly strict (https://data-blog.gbif.org/post/country-centroids/), although we chose a conservative distance for the Political centroid test (1 km).

Several reasons may explain the high proportion of records flagged as duplicates. First, the deposition of duplicates from the same specimen hosted at different collections is common practice, especially for plants, where a specimen duplication is easily feasible. Second, independent collections at similar localities may occur, in particular for local endemics. Third, low coordinate precision, for instance based on automated geo-referencing from locality descriptions, may lump records from nearby localities. Fourth, different data contributors might add the same record to GBIF, if their sources overlap, as can for instance be the case for the Barcode of Life and Plazi databases.

## Similarities and differences among taxa {-}
The number of records flagged by individual tests was similar across study groups, suggesting that similar problems might be relevant for collections of plants and animals. Therefore, the same filters can be used across taxonomic groups. Some notable exceptions stress the need to adapt each filter to the taxonomic study group to balance data quality and data availability. The high fraction of records filtered by the "basis of record filter" for *Tityus*, Dipsadidae, *Prosthechea* and *Tillandsia*, were mostly caused by a high number of records in these groups based on unknown collection methods, which might be caused by the contribution of specific datasets lacking this information for these groups. The high fraction of records flagged by the collection year test for Dispadidae was caused by a high collection effort already in the late 1880s and early 1900s, as can be expected for a charismatic group of reptiles, but also by 500 records dated to the year 1700. The latter records likely represent a data entry error: they are all contributed to GBIF from the same institution, and the institutions code for unavailable collection dates is 1700-01-01 - 2014-01-01, which has likely erroneously been converted to 1700. The high number of species flagged at capital coordinates and within urban areas for the plant groups Iridaceae and *Oocephalus* might be related to horticulture since at least some species in those groups are commonly cultivated as ornamentals.

@@

While our results suggest that large-scale species richness patterns and automated conservation assessments are largely robust to problematic occurrence records, they also stress the importance of (meta-)data exploration prior to most biodiversity analyses. Automated filtering can help to identify problematic records, but also highlight the necessity to customize tests and thresholds to the specific taxonomic groups and geographic area of interest. The putative problems we encountered point to the importance to train researchers and students to curate species occurrence datasets and to allow users to provide data aggregators such as GBIF with feedback for particular records, thus contributing to overall accuracy and precision of public biodiversity databases.
